Type-Safe Modular Parsing

### 2016-12-15 paper/sections/Overview.tex
RW

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Object Algebras~\cite{} enable us to solve this problem, and
offer high flexibility in the choice of operations to be performed
over the AST.

Instead of define the abstract syntax using Scala's case classes, we
can use \textit{Object Algebras} for more extensibility. Object
Algebras is a technique to solve the \textit{Expression Problem},
providing the possibility of extending both data variants and
operations over them. It will be further discussed in section
\ref{sec:algebrasandparsing}.

Using Object Algebras, the abstract syntax of the language of variables and applications is defined as below.

\begin{lstlisting}
trait Alg[E] {
  def Var(x: String): E
  def App(e1: E, e2: E): E
}
\end{lstlisting}

@@

\begin{lstlisting}
trait Print extends Alg[String] {
  def Var(x: String) = x
  def App(e1: String, e2: String) = "(" + e1 + " " + e2 + ")"
}
\end{lstlisting}

And also parser for it as below. Notice the parsing function \inlinecode{pVarApp} takes an argument of type \inlinecode{Alg[E]}, that means it accepts any instance of the algebra interface \inlinecode{Alg}.

\begin{lstlisting}
trait Parser[E] {
  lexical.delimiters += ("(", ")")

  val pVarApp: Alg[E] => Fix[PackratParser[E]] = alg => e =>
    e ~ e ^^ { case e1 ~ e2 => alg.App(e1, e2) } | ident ^^ alg.Var | "(" ~> e <~ ")"
}
\end{lstlisting}

We glue them together by defining an object. It actually becomes a \textit{language component} that represents a specific language feature.

\begin{lstlisting}
object VarApp {
  trait Alg[E] {..}
  trait Print extends Alg[String] {..}
  trait Parser[E] {..}
}
\end{lstlisting}

To use the parser, we must provide an algebra instance as the operation to construct the parsing result. As shown in the code below, we feed the pretty printing operation \inlinecode{VarApp.Print} to the generic \inlinecode{parse} function. It will print \inlinecode{(x y)} as the result.

\begin{lstlisting}
def parseWithAlg[E](inp: String)(alg: VarApp.Alg[E]): E = {
  val p = new VarApp.Parser[E] {}.pVarApp(alg)
  parse[E](p)(inp)
}

println(parseWithAlg("x y")(new VarApp.Print{}))
\end{lstlisting}

With Object Algebras, we can extend all the stuff, including abstract syntax, operations, and parsers. The code below extend the language to support lambda abstraction.

\begin{lstlisting}
object Lambda {
  trait Alg[E] extends VarApp.Alg[E] {
    def Lam(x: String, e: E): E
  }

  trait Print extends Alg[String] with VarApp.Print {
    def Lam(x: String, e: String) = "\\" + x + "." + e
  }

  trait Parser[E] extends VarApp.Parser[E] {
    lexical.delimiters += ("\\", ".")

    val pLam: Alg[E] => Fix[PackratParser[E]] = alg => e =>
      ("\\" ~> lcid) ~ ("." ~> e) ^^ { case x ~ e0 => alg.Lam(x, e0) }
    val pVarAppLam: Alg[E] => Fix[PackratParser[E]] =
      pVarApp | pLam
  }
}
\end{lstlisting}

For easily combining the parsers using Object Algebras as the representation of abstract syntax, we have a special combinator \huang{we need to rename our combinator} defined in our library.

In addition to make extending operations possible, another advantage of using Object Algebras is that it supports multiple sorts of syntax. For example, if our language has not only expressions but also types, we can easily distinguish them in the abstract syntax by an extra type parameter. This topic will be discussed in section \ref{subsec:differentsyntax}.

\begin{lstlisting}
trait TypedLambda[E, T] extends VarApp.Alg[E] {
  def Lam(x: String, t: T, e: E): E
  def IntType(): T
  def BoolType(): T
  def ArrowType(a: T, b: T): T
}
\end{lstlisting}

% \begin{itemize}
% \item Choosing the Parsing Technology
%     \begin{itemize}
%     \item Parser Generators : why not?
%         \begin{itemize}
%             \item Not type-safe
%             \item No modular type-checking
%             \item Not modular or no separate compilation (but we need to mention lots of work on extensible parsing here: example Language Workbenches; Rats)
%         \end{itemize}
%     \item Parser Combinators
%         \begin{itemize}
%         \item Backtracking parsers (Parsec)
%             \begin{itemize}
%             \item Need to remove left recursion (Problem: Transformation is not modular; if we do not know the full grammar then cannot be done)
%             \item Need try/backtracking (Problem: Since we do not know the full set of rules in a modular setting, we have to assume worst case scenario and add redundant backtracking. )
%             \item Mention some possible workarounds (there may be some but they still have issues)
%             \end{itemize}
%         \item (Non)-Backtracking/Packrat Parsers (Works well in a modular setting) (find a good name for this type of parsers?)
%         \end{itemize}
%     \end{itemize}
% \item Adapting Parser Combinators for modularity
%     \begin{itemize}
%     \item Using Object Algebras
%     \item Using Open Recursion
%     \item Using new modular parser combinators (Library: new alternative combinator, for example)
%     \end{itemize}
% \item Small example
%     \begin{itemize}
%     \item small example of a modular parser using our technique
%     \item show also the same example without modularity and write a detailed comparison
%     \item Lambda Calculus is a good candidate for the example
%     \item Show a small extension (adding plus and numeric literals)
%     \end{itemize}
% \end{itemize}

>>>>

Although delegation enables OO extensibility, the use of an OO class
hierarchy makes the addition of new operations over the AST
problematic. Object Algebras~\cite{} enable us to solve this problem, and
offer high flexibility in the choice of operations to be performed
over the AST. It also makes parsing with multiple sorts of syntax easier.

\paragraph{Parsing with Object Algebras} Using Object Algebras, the abstract syntax of the language of variables and applications is defined as below.

\lstinputlisting[linerange=8-11]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_ALG

@@

\lstinputlisting[linerange=15-18]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PRINT

And also parser for it as below. Notice the parsing function \inlinecode{pExpr} takes an argument of type \inlinecode{ExprAlg[E]}, that means it accepts any instance of the algebra interface \inlinecode{ExprAlg}.

\lstinputlisting[linerange=22-28]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_PARSER

To use the parser, we must provide an algebra instance as the operation to construct the parsing result. As shown in the code below, we feed the pretty printing operation \inlinecode{ExprPrint} to the generic \inlinecode{parse} function. It will print \inlinecode{(x y)} as the result.

\lstinputlisting[linerange=38-43]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_USE

\paragraph{Extensibility of Syntax} Following previous examples, the code below shows how to extend the language to support lambda abstractions.

\lstinputlisting[linerange=48-62]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT

For easily combining parsers using Object Algebras as ASTs, we have a special combinator \huang{we need to rename our combinator} defined in our library.

\paragraph{Extensibility of Operations} With Object Algebras, operations over ASTs can also be extended in a modular way. Here is an example of collecting free variables from an expression. We can use this new operation during parsing.

\lstinputlisting[linerange=75-81]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_EXT_OP

\paragraph{Multiple Sorts of Syntax} Another advantage of using Object Algebras is that it supports multiple sorts of syntax easily. In several cases, we want to divide the syntactic elements into some groups. For example, the grammar below has two sorts, which are expressions and types.

\begin{tabular}{m{0.4\linewidth}m{0.4\linewidth}}
\setlength{\grammarindent}{5em}
\begin{grammar}
<type> ::= `int' \alt <type> `->' <type>
\end{grammar}
&
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <ident> \alt <expr> <expr> \alt `\\' <ident> `.' <expr>
\end{grammar}
\end{tabular}

Using Object Algebras, we can easily distinguish them in an extensible way, just by adding an extra type parameter.

\lstinputlisting[linerange=66-70]{../Scala/Parser/src/PaperCode/Overview/OA.scala}% APPLY:linerange=OVERVIEW_OA_MULTI_SYNTAX
